The effects of multi-dimensional drought on land cover change and vegetation productivity in continental Chile

### 2024-03-19 article/paper_drought_Chile2023.qmd
update article

title: The effects of multidimensional drought on land cover change and vegetation productivity in continental Chile

@@

From 1960 to 2019, land use change has impacted around one-third of the Earth's surface, which is four times more than previously thought [@Winkler2021]. Multiple studies aim to analyze and forecast changes in land cover globally [@Winkler2021; @Song2018] and regionally [@Chamling2020; @Homer2020; @Yang2021]. Some others seek to analyze the impact of land cover change on climate conditions such as temperature and precipitation [@Luyssaert2014;@Pitman2012]. There is less research on the interaction between drought and land cover change [@Chen2022;@Akinyemi2021;@Peng2017]. @Peng2017 conducted a worldwide investigation utilizing net primary production to examine the spatial and temporal variations in vegetation productivity at global level. The study aimed to assess the influence of drought by comparing the twelve-month Standardized Precipitation Evapotranspiration Index (SPEI) and land cover change. According to their findings, drought is responsible for 37% of the decline in vegetation productivity, while water availability accounts for 55% of the variation. @Chen2022 studied the trend of vegetation greenness and productivity and its relation to meteorological drought (SPEI of twelve months in December) and soil moisture at the global level. The results showed lower correlations (<0.2) for both variables. @Akinyemi2021 evaluates drought trends and land cover change using vegetation indices in Botwsana in a semi-arid climate. These studies mostly looked at how changes in land cover and vegetation productivity are related to a single drought index (SPEI) over a single time period of 12 months. SPEI takes into account the combined effect of precipitation and AED as a water balance, but it does not allow us to know the contribution of each variable on its own. Some things worth investigating in terms of land cover change and vegetation productivity are: i) How do they respond to short- to long-term meteorological droughts? ii) How do they behave in humid and arid climatic zones regarding drought? And iii) What is the role of soil moisture? Likewise, there is a lack of understanding of how the alteration in water supply and demand is affecting land cover transformations.

Chile's diverse climatic and ecosystem types (@Beck2023;@Luebert2022) make it an ideal natural laboratory for studying climate and ecosystems. Additionally, the country has experienced severe drought conditions that have had significant effects on vegetation and water storage. Central Chile faced a persistent precipitation deficit between 2010 and 2022, defined as a megadrought [@Garreaud2017], which has impacted the Chilean ecosystem. This megadrought was defined by the Standardized Precipitation Index (SPI) of twelve months in December having values below one standard deviation. Some studies have addressed how this drought affects single ecosystems in terms of forest development [@Miranda2020;@Venegas2018], forest fire occurrence [@UrrutiaJalabert2018], and crop productivity [@Zambrano2023;@Zambrano2018;@Zambrano2016]. We found one study regarding land cover and drought in Chile. The study by @Fuentes2021  evaluates water scarcity and land cover change in Chile between 29° and 39° of south latitude. @Fuentes2021 used the SPEI of one month for evaluating drought, which led to misleading results. For example, they did not find a temporal trend in the SPEI but found a decreasing trend in water availability and an increase trend on AED, which in turn should have been capable of being captured with longer time scales of the SPEI. The term "megadrought" in Chile is used to describe a prolonged water shortage that lasts for several years, resulting in a permanent deficit that impacts the hydrological system [@Boisier2018]. Hence, it is imperative to assess temporal scales that take into account the cumnulative effect within some years. There is little knowledge about the relationship between drought and ecosystem in Chile; thus, it is important to understand in more detail how meteorological and soil moisture droughts influence ecosystem dynamics to inform adaptation options.

A detailed spatiotemporal assessment of the interaction of drought for short- to long-term and land cover change requires information on vegetation as well as weather variables such as precipitation, temperature, and soil moisture. Weather networks in Chile present some disadvantages, such as spatio-temporal gaps, a short history, and irregular quality, which make them difficult to represent the whole extent of the country spatially. In order to do this, we use reanalysis data from ERA5-Land [@MunozSabater2021] to create drought indices that consider AED, precipitation, and soil moisture over a range of time periods, from the short to the long term. Also, we use vegetation spectral information and annual land cover change from the Moderate-Resolution Imaging Spectroradiometer (MODIS). We expect to gain insight regarding the temporal evolution of water demand, water supply, and soil moisture, as well as the interaction with land cover change and vegetation productivity. Here, we analyze the multi-dimensional impacts of drought across ecosystems in continental Chile. More specifically, we aim to assess: i) temporal changes in land-use cover and the direction and magnitude of their relationships with drought indices for water demand and supply, soil moisture, and vegetation productivity; ii) short- to long-term temporal trends in multi-scalar drought indices; and iii) the relationship between vegetation productivity and drought indices for water demand and supply and soil moisture across Chilean ecosystems.

@@

Continental Chile has a diverse climate conditions with strong gradients from north to south and east to west [@Aceituno2021] (@fig-studyArea a), which determines its great ecosystem diversity (@Luebert2022) (@fig-studyArea c). The Andes Mountains are a main factor in climate latitudinal variation [@Garreaud2009]. In order to characterize the climate and ecosystem of Chile, we utilize the Köppen-Geiger classification system developed by @Beck2023 and the land cover data derived from the MODIS product for the period of 2001–2022, based on the International Geosphere-Biosphere Programme (IGBP) classification scheme proposed by @Friedl2019. "Norte Grande" and "Norte Chico" predominate in an arid desert climate with hot (Bwh) and cold (Bwk) temperatures. At the south of "Norte Chico," the climate changes to an arid steppe with cold temperatures (Bsk). In these two northern regions, the land is mostly bare, with a minor surface of vegetation types such as shrubland and grassland. In the zones "Centro" and the north half of "Sur," the main climate is Mediterranean, with warm to hot summers (Csa and Csb). Land cover in "Centro" comprises a significant amount of shrubland and savanna (50%), grassland (16%), forest (8%), and croplands (5%). An oceanic climate (Cfb) predominates in the south of "Sur" and the north of "Austral." Those zones are high in forest and grassland. The southern part of the country has a tundra climate, and in "Austral", it is a cold semi-arid area with an extended surface of grassland, forest, and, to a lesser extent, savanna.

@@

To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection 6.1 from MODIS. This product has a yearly frequency from 2001 to 2022. The IGBP defines 17 classes; from these, we regrouped into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to shrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies. Thus, we have a land cover raster time series with the ten classes for 2001 and 2023. Prior to using this class, we compared it with the types of land cover made by @Zhao2016, which is a more detailed land cover map of Chile with a 30 m spatial resolution for 2012–2013. It reached a global accuracy of ~0.82. The procedure of validation is demonstrated in Section S1 of the supplementary material.To derive a proxy for vegetation productivity, we used the MOD13A3 collection 6.1 product from MODIS [@Didan2015]. It provides vegetation indices (NDVI and EVI) at 1km of spatial resolution and monthly frequency. The MOD13A3 and MCD12Q1 were retrieved from the online Data Pool, courtesy of the NASA EOSDIS Land Processes Distributed Active Archive Center (LP DAAC), USGS Earth Resources Observation and Science (EROS) Center, Sioux Falls, South Dakota, https://lpdaa.usgs.gov/tools/data-pool/.

```{=latex}
\begin{table}[!ht]
\caption{Description of the earth observation data used }
\label{tab-desEOD}
\small
\centering
\begin{tabular}{p{0.13\textwidth}cp{0.3\textwidth}p{0.095\textwidth}ccc}
\hline
\multirow{1}{*}{\centering Product} & Sub-product & Variable & Spatial Resolution  & Period & Units & Short Name \\ 
\hline
\multirow{4}{*}{ERA5L} & ~ & Precipitation & \multirow{4}{*}{~0.1°} & \multirow{4}{*}{1981-2023} & mm & P \\ 
         &  & Maximum temperature & ~ & & $°C$ & $T_{max}$ \\ 
         &  & Minimum temperature & ~ & & $°C$ & $T_{min}$ \\ 
         &  & Volumetric Soil Water Content at 1m & ~ & & $m3/m3$ & SM \\ 
ERA5L* & & Atmospheric Evaporative Demand & 0.1° & 1981-2023 & mm & AED \\
        \multirow{2}{*}{MODIS} & MOD13A3.061 & Normalized Difference Vegetation Index & \multirow{2}{*}{~1 km} & 2000-2023 & ~ & NDVI \\ 
         & MCD12Q1.061 & land cover IGBP scheme & & 2001-2022 & ~ & land cover \\ 
\hline
\end{tabular}
{\raggedright *Derived from ERA5L with Eq. \ref{eq-AED}. \par}
\end{table}
```

For soil moisture, water supply, and water demand variables, we used ERA5L [@MunozSabater2021], a reanalysis dataset that provides the evolution of the meteorological and soil moisture variables since 1950. It has a spatial resolution of 0.1° (9 km), hourly frequency, and global coverage. We selected the variables for total precipitation, maximum and minimum temperature at 2 meters, and volumetric soil water layers between 0 and 100cm of depth (layer 1 to layer 3).

## Land cover change and trend

To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection 6.1 from MODIS. @Zambrano2018 and @Fuentes2021 have previously used this product for studies of drought and land cover. The MCD12Q1 has a yearly frequency from 2001 to 2022. The IGBP defines 17 classes; from these, we regrouped into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to shrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies. Thus, we have a land cover raster time series with the ten macroclasses for 2001 and 2023. We validate the land cover macroclasses regarding a highly detailed (30 m of spatial resolution) land cover map made for Chile by @Zhao2016 for 2013-2014. Our results showed a global accuracy of ~0.82 and a F1 score of ~0.66. Section S2 in the Supplementary Material shows the procedure for validation.

## Trend of drought and interaction to land cover change

@@

In order to compute the drought indices that uses water demand, it is necessary to first calculate the AED. To do this, we employed the Hargreaves technique [@Hargreaves1994;@Hargreaves1985], by applying the following equation:

@@

where $Ra$ $(MJ\,m^2\, day^{-1})$ is extraterrestrial radiation; $T$, $T_{max}$, and $T_{min}$ are mean, maximum, and minimum temperature $(°C)$. We calculate the centroid coordinates per pixel and use the latitude to estimate $Ra$.

We chose the method of Hargreaves to estimate AED because of its simplicity, which only requires temperatures and extrarrestrial radiation. Also, it has been recommended over other methods (e.g., Penman-Monteith) when the access to climatic variables is limited [@Vicente-Serrano2014].

@@

To derive the drought indices of water supply and demand, soil moisture, and vegetation we used the ERA5L dataset and the MODIS product, with a monthly frequency for 1981-2023 and 2000-2023, respectively.

@@

For the purpose of monitoring water supply drought, we used the well-known Standardized Precipitation Index (SPI), which the World Meteorological Organization (WMO) recommended. The SPI solely relies on precipitation data. Also, it has been used worldwide for the study of drought, including in Chile (@Garreaud2017; @Zambrano2017). The primary cause of drought is precipitation anomalies, and temperature usually makes it worse (Luo et al. 2017). Nowadays, there is an increase in attention toward using water demand separately to monitor droughts. (@Vicente-Serrano2020; @Noguera2022). Thus, to evaluate water demand, we chose the Evaporative Demand Drought Index (EDDI), developed by Hobbins et al. (016) and McEvoy et al. (2016), which is based on the AED. EDDI is currently used for monitoring drought in the United States (https://psl.noaa.gov/eddi/). In our case, we used only temperature for AED, a difference from the original formulation of EDDI, which also considered wind besides temperature. To consider the combined effect of water supply and demand, we selected the SPEI, which corresponds to a balance between precipitation and AED. @Vicente-Serrano2010 proposed the SPEI, and it has improved the SPI by incorporating temperature for drought monitoring. For SPEI, an auxiliary variable D = P-AED is calculated. Soil moisture is the main driver of vegetation productivity, particularly in semi-arid regions (@Li2022). Hence, for soil water drought, we used the SSI (Standardized Soil Moisture Index) (Hao and AghaKouchak 2013; A. AghaKouchak 2014) which is a multi-scale index similar to SPI, SPEI, and EDDI. In our case, for the SSI, we used the average soil moisture from ERA5L at 1m depth. Finally, for the proxy of productivity, we used the zcNDVI proposed by Zambrano et al. (2018) which will be derived from the NDVI retrieved from MOD13A1.

@@

### Intercation drought indices and land cover change

@@

@@

# Land cover change and trend

## Trend of drought and interaction to land cover change

@@

@@

According to Table \ref{tab-landcoverTrendRF}, barren land is the type of land cover that is most affected by the trend of drought indices that reach a rsq = 0.4. This is attributed to the fact that it is linked to short-term deficits in precipitation (SPI-6 and SPEI-6) and long-term AED (EDDI-12 and EDDI-24). The trend in drought indices accounts for about a quarter of the change in grassland, savanna, and shrubland types. These types are most affected by the increase in short-term water demand (EDDI-6). The changes in forests are strongly impacted by the increase in long-term water demand (EDDI-36). The trend in croplands cannot be attributed to variations in water demand, water supply, soil moisture, or vegetation productivity.

@@

```{r fig-TrendsLandDrought}
#| echo: false
#| fig-pos: "!ht"
#| fig-cap: "Relationship between the trend in land cover change and the trend in drought indices for the five macrozones. Vertical panels correspond to 1, 3, 6, 12, 24, and 36 months of the time scale by drought index. Horizontal panels show each drought index"
#| out-width: "\\textwidth"
knitr::include_graphics('../output/figs/points_landcover_drought_indices_trend_and_time_scale.png')
```

We look at the relationship between drought index and land cover change in @fig-TrendsLandDrought by comparing the trends in land cover change (in terms of the total surface area per land cover type and macrozone) and drought indexes. @fig-TrendsLandDrought shows that the negative trends in cropland (-0.029) and savanna (-0.026) in “Norte Chico” are the highest and are associated with an increase trend in EDDI and a decreasing trend in SPI, SPEI, and SSI. On the contrary, the shrubland in “Centro” has an increase (0.04) linked to a decrease in SPI, SPEI, and SSI and to an increase in EDDI. In "Austral,” the positive trend in shrubland fits with the positive trend in all the drought indices. The rest of the land cover types in the macrozones show weak associations with water supply, water demand, or soil moisture drought indices.

@@

In @fig-zcNDVI_var it is showed the spatial map of trends in zcNDVI ([fig-zcNDVI_var]a) and the temporal variation of zcNDVI within the aggregated macrozones ([fig-zcNDVI_var]b). In "Norte Grande," vegetation productivity, as per the z-index, exhibits a yearly increase of 0.02 with respect to grassland and shrubland categories. There is a negative trend in "Norte Chico" with -0.04 and "Centro" with -0.02 per decade. In the "Norte Chico," savanna (-0.05) has the lowest trend, and the rest of the types are around -0.04. In "Centro," shrubland reaches -0.06, grassland -0.05, and croplands and savanna -0.01 per decade. This could be associated either with a reduction in vegetation surface, a decrease in biomass, or browning [@Miranda2023]. Vegetation reached its lowest values since the year 2019, with an extreme condition in early 2020 and 2022 in the "Norte Chico" and "Centro". The "Sur" and "Austral" show a positive trend of around 0.016 per decade (@fig-zcNDVI_var). Despite the croplands suffering from drought just as badly as the native vegetation in "Norte Chico," the savanna and shurbland appears to be the region most affected by a negative trend in vegetation productivity.

>>>>

title: The effects of multi-dimensional drought on land cover change and vegetation productivity in continental Chile

@@

From 1960 to 2019, land use change has impacted around one-third of the Earth's surface, which is four times more than previously thought [@Winkler2021]. Multiple studies aim to analyze and forecast changes in land cover globally [@Winkler2021; @Song2018] and regionally [@Chamling2020; @Homer2020; @Yang2021]. Some others seek to analyze the impact of land cover change on climate conditions such as temperature and precipitation [@Luyssaert2014;@Pitman2012]. There is less research on the interaction between drought and land cover change [@Chen2022;@Akinyemi2021;@Peng2017]. @Peng2017 conducted a worldwide investigation utilizing net primary production to examine the spatial and temporal variations in vegetation productivity at global level. The study aimed to assess the influence of drought by comparing the twelve-month Standardized Precipitation Evapotranspiration Index (SPEI) and land cover change. According to their findings, drought is responsible for 37% of the decline in vegetation productivity, while water availability accounts for 55% of the variation. @Chen2022 studied the trend of vegetation greenness and productivity and its relation to meteorological drought (SPEI of twelve months in December) and soil moisture at the global level. The results showed lower correlations (<0.2) for both variables. @Akinyemi2021 evaluates drought trends and land cover change using vegetation indices in Botwsana in a semi-arid climate. These studies mostly looked at how changes in land cover and vegetation productivity are related to a single drought index (SPEI) over a single time period of 12 months. SPEI takes into account the combined effect of precipitation and AED as a water balance, but it does not allow us to know the contribution of each variable on its own. Some things worth investigating in terms of land cover change and vegetation productivity are: i) How do they respond to short- to long-term meteorological and soil moisture droughts? ii) How is the drought impacting land cover changes? And iii) How do they behave in humid and arid climatic zones regarding drought? Likewise, there is a lack of understanding of how the alteration in water supply and demand is affecting land cover transformations.

For monitoring drought, the World Meteorological Organization recommends the SPI (Standardized Precipitation Index) [@WMO2012]. The SPI is a multi-scalar drought index that only uses precipitation to assess short- to long-term droughts. The primary cause of drought is precipitation anomalies, and temperature usually makes it worse [@Luo2017]. Nowadays, there is an increase in attention toward using AED separately to monitor droughts [@Vicente-Serrano2020]. One reason is due to its attribution to increasing flash droughts in water-limited regions [@Noguera2022]. @Vicente-Serrano2010 proposed the Standardized Precipitation Evapotranspiration Index (SPEI), which incorporated the temperature effect by subtracting AED from precipitation. SPEI allows for analysis of the combined effect of precipitation and AED. Since its formulation, it has been used worldwide for the study and monitoring of drought [@Gebrechorkos2023;@Liu2021]. @Hobbins2016 and @McEvoy2016 developed the Evaporative Demand Drought Index (EDDI) to monitor droughts solely using the AED, and it has proven effective in monitoring flash droughts [@Li2024;@Ford2023]. For soil moisture, several drought indices exist, such as the Soil Moisture Deficit Index (SDMI) [@Narasimhan2005] and the Soil Moisture Agricultural Drought Index (SMADI) [@Souza2021]. @Hao2013 and @AghaKouchak2014 proposed the Standardized Soil Moisture Index (SSI), which has a similar formulation as the SPI, SPEI, and EDDI. Thus, there are plenty of drought indices that allow for a comprehensive assessment of drought from short- to long-term scales and consider variables from the earth's water balance on their own (e.g., precipitation, AED, soil moisture). Using this information, we can advance our understanding of the impact of drought on ecosystems.

Chile's diverse climatic and ecosystem types [@Beck2023;@Luebert2022] make it an ideal natural laboratory for studying climate and ecosystems. Additionally, the country has experienced severe drought conditions that have had significant effects on vegetation and water storage. North-central Chile has faced a persistent precipitation deficit since 2010, defined as a mega drought. [@Garreaud2017], which has impacted the Chilean ecosystem. This megadrought was defined by the Standardized Precipitation Index (SPI) of twelve months in December having values below one standard deviation. Some studies have addressed how this drought affects single ecosystems in terms of forest development [@Miranda2020;@Venegas2018], forest fire occurrence [@UrrutiaJalabert2018], and crop productivity [@Zambrano2023;@Zambrano2018;@Zambrano2016]. We found one study regarding land cover and drought in Chile. The study by @Fuentes2021  evaluates water scarcity and land cover change in Chile between 29° and 39° of south latitude. @Fuentes2021 used the SPEI of one month for evaluating drought, which led to misleading results. For example, they did not find a temporal trend in the SPEI but found a decreasing trend in water availability and an increase trend on AED, which in turn should have been capable of being captured with longer time scales of the SPEI. The term "megadrought" in Chile is used to describe a prolonged water shortage that lasts for several years, resulting in a permanent deficit that impacts the hydrological system [@Boisier2018]. Therefore, it is crucial to evaluate temporal scales that consider the cumulative impact over a period of several years. There is limited understanding of the correlation between drought and the environment in Chile. Therefore, it is crucial to have a deeper understanding of how meteorological and soil moisture droughts impact the dynamics of the environment in order to make informed decisions on how to adapt.

<!-- A detailed spatiotemporal assessment of the interaction of drought for short- to long-term and land cover change requires information on vegetation as well as weather variables such as precipitation, temperature, and soil moisture. Weather networks in Chile present some disadvantages, such as spatio-temporal gaps, a short history, and irregular quality, which make them difficult to represent the whole extent of the country spatially. In order to do this, we use reanalysis data from ERA5-Land [@MunozSabater2021] to create drought indices that consider AED, precipitation, and soil moisture over a range of time periods, from the short to the long term. Also, we use vegetation spectral information and annual land cover change from the Moderate-Resolution Imaging Spectroradiometer (MODIS). We expect to gain insight regarding the temporal evolution of water demand, water supply, and soil moisture, as well as the interaction with land cover change and vegetation productivity.  -->

Here, we analyze the multi-dimensional impacts of drought across ecosystems in continental Chile. More specifically, we aim to assess: i) short- to long-term temporal trends in multi-scalar drought indices; ii) temporal changes in land-use cover and the direction and magnitude of their relationships with trends in drought indices; and iii) the trend in vegetation productivity and its relationship with drought indices across Chilean ecosystems.

@@

Continental Chile has diverse climate conditions with strong gradients from north to south and east to west [@Aceituno2021] (@fig-studyArea a), which determines its great ecosystem diversity [@Luebert2022] (@fig-studyArea c). The Andes Mountains are a main factor in climate latitudinal variation [@Garreaud2009]. "Norte Grande" and "Norte Chico" predominate in an arid desert climate with hot (Bwh) and cold (Bwk) temperatures. At the south of "Norte Chico," the climate changes to an arid steppe with cold temperatures (Bsk). In these two northern regions, the land is mostly bare, with a small surface of vegetation types such as shrubland and grassland. In the zones "Centro" and the north half of "Sur," the main climate is Mediterranean, with warm to hot summers (Csa and Csb). Land cover in "Centro" comprises a significant amount of shrubland and savanna (50%), grassland (16%), forest (8%), and croplands (5%). An oceanic climate (Cfb) predominates in the south of "Sur" and the north of "Austral." Those zones are high in forest and grassland. The southern part of the country has a tundra climate, and in "Austral," it is a cold semi-arid area with an extended surface of grassland, forest, and, to a lesser extent, savanna.

@@

To analyze land cover change, we use the classification scheme by the IGBP (International Geosphere-Biosphere Programme) from the product MCD12Q1 collection 6.1 from MODIS. To derive a proxy for vegetation productivity, we used the Normalized Difference Vegetation Index (NDVI) from the product MOD13A3 collection 6.1 from MODIS [@Didan2015]. MOD13A3 provides vegetation indices at 1km of spatial resolution and monthly frequency. The NASA EOSDIS Land Processes Distributed Active Archive Center (LP DAAC), USGS Earth Resources Observation and Science (EROS) Center, Sioux Falls, South Dakota, provided the MOD13A3 and MCD12Q1 from the online Data Pool, accessible at https://lpdaa.usgs.gov/tools/data-pool/.

```{=latex}
\begin{table}[!ht]
\caption{Description of the satellite and reanalysis data used}
\label{tab-desEOD}
\small
\centering
\begin{tabular}{p{0.13\textwidth}cp{0.3\textwidth}p{0.095\textwidth}ccc}
\hline
\multirow{1}{*}{\centering Product} & Sub-product & Variable & Spatial Resolution  & Period & Units & Short Name \\ 
\hline
\multirow{4}{*}{ERA5L} & ~ & Precipitation & \multirow{4}{*}{~0.1°} & \multirow{4}{*}{1981-2023} & mm & P \\ 
         &  & Maximum temperature & ~ & & $°C$ & $T_{max}$ \\ 
         &  & Minimum temperature & ~ & & $°C$ & $T_{min}$ \\ 
         &  & Volumetric Soil Water Content at 1m & ~ & & $m3/m3$ & SM \\ 
ERA5L* & & Atmospheric Evaporative Demand & 0.1° & 1981-2023 & mm & AED \\
        \multirow{2}{*}{MODIS} & MOD13A3.061 & Normalized Difference Vegetation Index & \multirow{2}{*}{~1 km} & 2000-2023 & ~ & NDVI \\ 
         & MCD12Q1.061 & land cover IGBP scheme & & 2001-2022 & ~ & land cover \\ 
\hline
\end{tabular}
{\raggedright *Calculated from maximum and minimum temperatures derived from ERA5L with Eq. \ref{eq-AED}. \par}
\end{table}
```

For soil moisture, water supply, and water demand variables, we used ERA5L (ECMWF Reanalysis version 5 over land) [@MunozSabater2021], a reanalysis dataset that provides the evolution of atmospheric and land variables since 1950. It has a spatial resolution of 0.1° (9 km), hourly frequency, and global coverage. We selected the variables for total precipitation, maximum and minimum temperature at 2 meters, and volumetric soil water layers between 0 and 100cm of depth (layer 1 to layer 3). Table \ref{tab-desEOD} shows a summary of the data and its main characteristics.

## Trend of short- to long-term drought

@@

In order to compute the drought indices that use water demand, it is necessary to first calculate the AED. To do this, we employed the Hargreaves method [@Hargreaves1994;@Hargreaves1985] by applying the following equation:

@@

where $Ra$ $(MJ\,m^2\, day^{-1})$ is extraterrestrial radiation; $T$, $T_{max}$, and $T_{min}$ are mean, maximum, and minimum temperature $(°C)$ at 2m. For calculating $Ra$ we used the coordinate of the latitud of the centroid of each pixel. We chose the method of Hargreaves to estimate AED because of its simplicity, which only requires temperatures and extrarrestrial radiation. Also, it has been recommended over other methods (e.g., Penman-Monteith) when the access to climatic variables is limited [@Vicente-Serrano2014].

@@

To derive the drought indices of water supply and demand, soil moisture, and vegetation (i.e., the proxy of productivity), we used the ERA5L dataset and the MODIS product, with a monthly frequency for 1981–2023 and 2000–2023, respectively.

@@

For the purpose of monitoring water supply drought, we used the well-known Standardized Precipitation Index (SPI), which the World Meteorological Organization (WMO) recommended. The SPI solely relies on precipitation data. Also, it has been used worldwide for the study of drought, including in Chile (@Garreaud2017; @Zambrano2017). The primary cause of drought is precipitation anomalies, and temperature usually makes it worse [@Luo2017]. Nowadays, there is an increase in attention toward using water demand separately to monitor droughts. (@Vicente-Serrano2020; @Noguera2022). Thus, to evaluate water demand, we chose the Evaporative Demand Drought Index (EDDI), developed by @Hobbins2016 and @McEvoy2016, which is based on the AED. EDDI is currently used for monitoring drought in the United States (https://psl.noaa.gov/eddi/). In our case, we used only temperature for AED, a difference from the original formulation of EDDI, which also considered wind besides temperature. To consider the combined effect of water supply and demand, we selected the SPEI, which corresponds to a balance between precipitation and AED. @Vicente-Serrano2010 proposed the SPEI, and it has improved the SPI by incorporating temperature for drought monitoring. For SPEI, an auxiliary variable D = P-AED is calculated. Soil moisture is the main driver of vegetation productivity, particularly in semi-arid regions (@Li2022). Hence, for soil water drought, we used the SSI (Standardized Soil Moisture Index) [@Hao2013;@AghaKouchak2014] which is a multi-scale index similar to SPI, SPEI, and EDDI. In our case, for the SSI, we used the average soil moisture from ERA5L at 1m depth. Finally, for the proxy of productivity, we used the zcNDVI proposed by @Zambrano2018 which will be derived from the NDVI retrieved from MOD13A1.

@@

## Interaction of land cover and drought

### Land cover change

To analyze the land cover change, we use the IGBP scheme from the MCD12Q1 collection 6.1 from MODIS. @Zambrano2018 and @Fuentes2021 have previously used this product for studies of drought and land cover in Chile. The MCD12Q1 has a yearly frequency from 2001 to 2022. The IGBP defines 17 classes; from these, we regrouped into ten macroclasses, as follows: classes 1-4 to forest, 5-7 to shrublands, 8-9 to savannas, 10 as grasslands, 11 as wetlands, 12 and 14 to croplands, 13 as urban, 15 as snow and ice, 16 as barren, and 17 to water bodies. Thus, we have a land cover raster time series with the ten macroclasses for 2001 and 2023. We validate the land cover macroclasses regarding a highly detailed (30 m of spatial resolution) land cover map made for Chile by @Zhao2016 for 2013-2014. Our results showed a global accuracy of ~0.82 and a F1 score of ~0.66. Section S2 in the Supplementary Material shows the procedure for validation.

### Relationship between land cover and drought trends

@@

## Drought impacts on vegetation productivity

@@

## Trend of short- to long-term drought

@@

### Interaction of land cover and drought

## Land cover change

@@

According to Table \ref{tab-landcoverTrendRF}, the trends in drought indices reach an r-squared between 0.32 and 0.39 for the types of forest, grassland, savanna, shrubland, and barren land. It is more likely that short- and medium-term increases in AED (EDDI-6 and EDDI-12) and short-term precipitation deficits (SPI-6 and SPEI-6) contributed to changes in grassland and bare land. The short-term increase of AED (EDDI-3 and EDDI-6) and the longer duration of the precipitation deficit (SPI-24, SPI-36, and SPEI-36) most likely contribute to the changes in shrubland. The changes in savanna are associated with a short- and long-term increase in AED and a three-year precipitation deficit (SPI-36). The increase in cumulative AED from 12 to 36 months is the strongest associated variable that contributes to changes in forests, followed by the reduction of soil moisture over six and 36 months. More details about the results of the variable importance can be found in the supplementary material in Section S3.

@@

```{r fig-TrendsLandDrought}
#| echo: false
#| fig-pos: "!ht"
#| fig-cap: "Relationship between the trend in land cover change (y-axis) and the trend in drought indices (x-axis) for the five macrozones. Vertical panels correspond to 1, 3, 6, 12, 24, and 36 months of the time scale by drought index. Horizontal panels show each drought index"
#| out-width: "\\textwidth"
knitr::include_graphics('../output/figs/points_landcover_drought_indices_trend_and_time_scale.png')
```

We study the connection between the SPI, EDDI, and SSI drought indices and changes in land cover in @fig-TrendsLandDrought. To do this, we compare the relative changes in land cover (in terms of the total surface area per land cover type and macrozone) over six and thirty-six months. @fig-TrendsLandDrought shows that the forest in the "Sur," shrubland and grassland in "Centro," barren land in "Norte Chico," and savanna in "Austral" showed an increase in the surface of landcover associated with an increase in EDDI. Savanna in "Centro," "Sur," and "Norte Chico" decreases with the increase in EDDI. The SPI and SSI showed similar behavior regarding the trend in land cover type. A decrease in SPI and SSI is associated with an increase in the surface in shurubland and grassland in "Centro," forest in "Sur," and barren land in "Norte Chico," as well as a decrease trend in savanna in "Norte Chico," "Centro," and "Sur."

@@

In @fig-zcNDVI_var it is showed the spatial map of trends in zcNDVI ([@fig-zcNDVI_var]a) and the temporal variation of zcNDVI within the aggregated macrozones ([@fig-zcNDVI_var]b). In "Norte Grande," vegetation productivity, as per the z-index, exhibits a yearly increase of 0.02 with respect to grassland and shrubland categories. There is a negative trend in "Norte Chico" with -0.04 and "Centro" with -0.02 per decade. In the "Norte Chico," savanna (-0.05) has the lowest trend, and the rest of the types are around -0.04. In "Centro," shrubland reaches -0.06, grassland -0.05, and croplands and savanna -0.01 per decade. This could be associated either with a reduction in vegetation surface, a decrease in biomass, or browning [@Miranda2023]. Vegetation reached its lowest values since the year 2019, with an extreme condition in early 2020 and 2022 in the "Norte Chico" and "Centro". The "Sur" and "Austral" show a positive trend of around 0.016 per decade (@fig-zcNDVI_var). Despite the croplands suffering from drought just as badly as the native vegetation in "Norte Chico," the savanna and shurbland appears to be the region most affected by a negative trend in vegetation productivity.
